---
title: "Data Wrangling"
author: "Umbu Domu Mahani"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Knitting options

library( dplyr )    # data wrangling
library( pander )   # formatting output
library( ggmap )    # grab map tiles
library( viridis )  # color pallette for maps
library( ggplot2 )  # fancy graphics
library( ggthemes ) # fancy themes for ggplots

```

<br>

# Source Data

The following analysis uses data comprised of traffic accidents in the city of 
Tempe, Arizona. They constitute 28,470 observations and 32 variables. You can 
find more details in the dataset documentation at the 
[**Tempe Open Data Portal**](https://data.tempe.gov/datasets/tempegov::1-08-crash-data-report-detail/about). 

```{r include = FALSE}

# READ IN DATA

url <- paste0("https://github.com/DS4PS/Data-Science-Class/blob",
              "/master/DATA/TempeTrafficAccidents.rds?raw=true")

dat <- readRDS(gzcon(url(url)))     # Method per instructions

```

<br>

## Preview

We can explore the first six observations in the following table using function `head()`.

```{r}

head(dat)

```

```{r}

glimpse(dat)

```

<br>
<br>

# Preprocessing

All data preprocessing are available in **Appendix A: Data Preprocessing**.

```{r include = FALSE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     labels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>
<br>

# Part I: Summaries

The following provides brief exploratory summaries on key data points.

<br>

## 1. Mondays Accidents

**Question:** *How many accidents happened on Mondays?*

**Note:** *This solution has been completed for you.*

<br>

```{r}

sum(dat$day == "Mon", 
    na.rm = TRUE)

```

<br>

**Answer:** **4,094** accidents occurred on Mondays.

<br>

## 2. Monday Accidents (%)

**Question:** *What proportion of accidents each week occur on Monday?*

<br>

```{r}

prop_mon <- mean(dat$day == "Mon", na.rm = TRUE) * 100
prop_mon

```

<br>

**Answer:** **14.38%** of all accidents occur on Mondays.

<br>

## 3. Harmful Monday Accidents (%)

**Question:** *What proportion of accidents on Mondays result in harm?*

**Note:** *"Harm" is defined as any accident that causes at least one injury or* 
*fatality, a.k.a. a "casualty".*

<br>

```{r}

dat$harm <- (dat$Totalinjuries > 0 | dat$Totalfatalities > 0) 

harm_prop <- mean(dat$harm[dat$day == "Mon"], na.rm = TRUE) * 100
harm_prop

```

<br>

**Answer:** **29.84%** of all Monday accidents have at least one casualty.

<br>

## 4. Most Common Accidents

**Question:** *What is the most common type of accident (Collisionmanner) that occurs on Mondays?*

<br>

```{r}

dat %>%
  filter(day == "Mon") %>%
  count(Collisionmanner) %>%
  arrange(desc(n)) %>%
  slice(1)

```

<br>

**Answer:** **Rear End** collisions are the most common accident types.

<br>

## 5. Differences in Accidents

**Question:** *Are there differences in the proportion of accidents that result in harm each day of the week?*

<br>

```{r}

dat %>%
  group_by(day) %>%
  summarise(
    harm_prop = mean(harm, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
arrange(desc(harm_prop))
```

<br>

**Answer:** Significantly more accidents occur **on Thursday (32.69%)**.

<br>

## 6. Accident Stats by Weekday

**Instructions:** Create a table that reports the following for each day of the week.

* Total accidents
* Total injuries
* Total fatalities
* Proportion of harmful accidents (i.e. casualties)

<br>

```{r}

dat %>%
  group_by(day) %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(n = n(),
            injuries = sum(Totalinjuries),
            fatalities = sum(Totalfatalities),
            harm.rate = mean(harm_acc))

```

<br>
<br>

# Part II: Age Groups

The following provides summaries of accidents by age groups.

<br>

## 1. Accidents by Hour & Age

**Instructions:** Create a table of counts of accidents by time of day (`hour12`) and age of driver (`age`).

**Question:** *Which age group has the largest number of accidents at 7AM?*

<br>

```{r}

dat %>%
  filter(hour12 == "7 AM") %>%
  count(Age_Drv1) %>%
  arrange(desc(n))

```

<br>

**Answer:** Drivers of **21** have the greatest number of accidents from 7:00 to 7:59 AM.

<br>

## 2. Accidents by Hour & Age (%)

**Instructions:** Create a new table of time of day and age group that reports the proportion of accidents at "7 AM" (`hour12`) for each age group (`age`). *The proportions within each age group should sum to one.*

<br>
<br>

**Objective:** The table should contain the following columns and initial values:

```{r echo = FALSE}

dat %>% 
  # 1. Calculate the total number of accidents per time (hour12) and age group combination
  group_by(hour12, age) %>% 
  summarize(n = n(), .groups = "drop") %>% 
  
  # 2. Compute the sub-total of accidents for each group (by age and by hour separately)
  group_by(age) %>% 
  mutate(n.age = sum(n)) %>% 
  ungroup() %>% 
  group_by(hour12) %>% 
  mutate(n.hour = sum(n)) %>% 
  ungroup() %>% 
  
  # 3. Calculate global proportion (p), proportion within age group (p.age), and proportion within hour group (p.hour)
  mutate(
    total_acc = sum(n),
    p = round(n / total_acc, 2),
    p.age = round(n / n.age, 2),
    p.hour = round(n / n.hour, 2)
  ) %>% 
  
  # 4. Filter specifically for the 7 AM time slot, remove missing values, select target columns, and sort
  filter(hour12 == "7 AM" & !is.na(age)) %>% 
  select(age, hour12, n, n.age, n.hour, p, p.age, p.hour) %>% 
  arrange(age) %>% 
  pander()

```

<br>
<br>

**Solution:** *The solution has been partially provided. Use the hints provided to create the table.*

```{r}

dat %>% 
  group_by(hour12, age) %>% 
  summarize(n = n()) %>%      # Partial solution
  group_by(age) %>%           # Requires two new variables in 'mutate()'
  mutate()                    # Lastly, requires function 'filter()

```

<br>
<br>

# Part III: Rates of Harm

The following reports the accidents, casualties, proportion, and average 
casualties per harmful accident.

<br>

## 1. Accidents by Hour

**Instructions:** Visualize total accidents by time of day (`hour`).

**Note:** *This solution has been completed for you.*

<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarize(n = n()) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour",
       ylab = "Total Number of Accidents",
       main = "Total Number of Accidents by Time of Day")

```

<br>

## 2. Total Casualties by Hour

**Instructions:** Visualize total injuries and fatalities by time of day (`hour`).

<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarise(
    total_injuries = sum(Totalinjuries, na.rm = TRUE),
    total_fatalities = sum(Totalfatalities, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = c(total_injuries, total_fatalities),
                      names_to = "type",
                      values_to = "count") %>%
  ggplot(aes(x = hour, y = count, fill = type)) +
  geom_col(position = "dodge") +
  labs(title = "Total Casualties by Hour",
       x = "Hour of Day",
       y = "Count",
       fill = "Casualty Type") +
  theme_minimal()

```

<br>

## 3. Accidents with Casualties (%)

**Instructions:** Visualize the proportion of harmful accidents out of all accidents by time of day (`hour`).

<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarise(
    total_accidents = n(),
    harmful_accidents = sum(harm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(harm_rate = harmful_accidents / total_accidents * 100) %>%
  ggplot(aes(x = as.numeric(hour), y = harm_rate)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkred", size = 2) +
  labs(title = "Proportion of Harmful Accidents by Hour",
       x = "Hour of Day",
       y = "Harmful Accident Rate (%)") +
  theme_minimal()

```

<br>

## 4. Avg. Harm by Harmful Accident

**Instructions:** Visualize average injuries or fatalities per harmful collision.

<br>

```{r}

dat %>%
  filter(harm == 1) %>%   # only harmful accidents
  group_by(hour) %>%
  summarise(
    avg_injuries = mean(Totalinjuries, na.rm = TRUE),
    avg_fatalities = mean(Totalfatalities, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = c(avg_injuries, avg_fatalities),
                      names_to = "type",
                      values_to = "average") %>%
  ggplot(aes(x = hour, y = average, color = type, group = type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Average Harm per Harmful Accident by Hour",
       x = "Hour of Day",
       y = "Average per Harmful Accident",
       color = "Harm Type") +
  theme_minimal()

```

<br>

# Appendix

The following code is used for preprocessing tasks in the above solutions.

<br>

## A: Data Preprocessing

These expressions transform, format, and create myriad new variables used in this report.

<br>

```{r eval = FALSE, echo = TRUE, include = TRUE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     levels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>
<br>
